MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-26 main/MSstatsPTM_Main.tex
Small suggestions for results

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance (Anova)\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

@@

We compared the performance of  $MSstatsPTM$ to that of ANOVA and of Limma. Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons. $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation. All the evaluations were done at the FDR-adjusted p-value cutoff of \todo{typically adjusted p-values are denoted $q$, and $\alpha$ is left for the unadjusted p-values. E.f., Eq. 12 uses $q$. Could you check the entire manuscript for consistency} $\alpha=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In clean simulations, MSstatsPTM corrected for high FDR of differentially abundant PTM}

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and Limma performed similarly. As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of Anova lagged behind, especially in cases with higher simulated variance. This was due to Anova's more simplistic variance estimation. In contrast, $MSstatsPTM$ and Limma implement methods to provide better estimates of the variance, resulting in improved power. Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and Anova. As discussed previously, these methods are not directly applicable to experiments with additional variance components.

@@

In simulations with missing values and few features, $MSstatsPTM$ outperformed Anova and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy  well, even when the number of replicates were low. $MSstatsPTM$ can handle low feature counts better than the existing methods due to using Tukey's Median Polish for summarization. The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.

Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and Anova's IQR. This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the existing method's distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and Anova's IQR. While Limma and Anova overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate. The proposed method's estimation was more consistent, with a smaller error. As discussed in the previous section, the summarization methods of the proposed methods excel when the number of observations are low. This was the case here, with each PTM only averaging $1.37$ features. The existing approaches used a non-robust summarization method, that was very sensitive to few observations, causing their fold change estimation to be more inconsistent, with a larger error compared to the proposed method.

@@

We proposed a general statistical modeling framework and implementation for PTM characterization. The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions. The framework is general and is applicable to a variety of experimental designs. It outperforms the ad-hoc methods underlying Anova and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods. The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance. The implementation allows for straightforward application of the methods discussed and allows for reproducible experimental analysis.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{Dataset 1 \& 2: Computer simulation. a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates. b) The advantage of using the proposed approach was apparent when including limited observations and missing values. Looking at accuracy, the proposed method outperformed Anova and Limma in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and Anova models was wider than the proposed method. b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}

>>>>

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance (ANOVA)\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Evaluation criteria}

@@

We compared the performance of  $MSstatsPTM$ to that of ANOVA and of Limma. Since IsobarPTM is only applicable to experiments with TMT labeling, it could not be applied to the datasets with known ground truth in this manuscript, and was excluded from the comparisons. $MSstatsPTM$ was used without imputing missing values, and without Empirical Bayes moderation. All the evaluations were done at the FDR-adjusted p-value cutoff of \todo{typically adjusted p-values are denoted $q$, and $\alpha$ is left for the unadjusted p-values. E.g., Eq. 12 uses $q$. Could you check the entire manuscript for consistency} $\alpha=.05$. More details are in {\bf Supplementary Sec. \ref{supp-sec:experiments}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In clean simulations, MSstatsPTM corrected the excess FDR in unadjusted differentially abundant PTM}

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and Limma performed similarly. As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates \todo{Hmm... figure 4a doesn't show the sample size. Did you mean Fig 4b? Then it should be in the next paragraph. If you refer to supplementary figures, say so first}. In contrast, the performance of ANOVA lagged behind, especially in cases with higher simulated variance. This was due to ANOVA's more simplistic variance estimation. In contrast, $MSstatsPTM$ and Limma implement methods to provide better estimates of the variance, resulting in improved power. Finally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and ANOVA. As discussed previously, these methods are not directly applicable to experiments with additional variance components.

@@

In simulations with missing values and few features, $MSstatsPTM$ outperformed ANOVA and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy  well, even when the number of replicates were low. $MSstatsPTM$ can handle low feature counts better than the existing methods due to using Tukey's Median Polish for summarization. The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.

Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 32.5\% smaller than Limma and ANOVA's IQR. This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the existing method's distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 32.86\% smaller than Limma and ANOVA's IQR. While Limma and ANOVA overall estimated the correct fold change, they were inconsistent in their estimation, resulting in a higher error rate. The proposed method's estimation was more consistent, with a smaller error. As discussed in the previous section, the summarization methods of the proposed methods excel when the number of observations are low. This was the case here, with each PTM only averaging $1.37$ features. The existing approaches used a non-robust summarization method, that was very sensitive to few observations, causing their fold change estimation to be more inconsistent, with a larger error compared to the proposed method.

@@

We proposed a general statistical modeling framework and implementation for PTM characterization. The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple confounded sources, frequent missing data, and associated uncertainty in the conclusions. The framework is general and is applicable to a variety of experimental designs. It outperforms the ad-hoc methods underlying ANOVA and Limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods. The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance. The implementation allows for straightforward application of the methods discussed and allows for reproducible experimental analysis.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{\todo{In the figure titles and in the caption, explain which simulation is 'clean' and which is 'noisy'. Use the same names for the datasets as in the Datasets section throughout. ANOVA is an abbreviation and needs all capital letters - could you check throughout in the figures} Dataset 1 \& 2: Computer simulation. a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates. b) The advantage of using the proposed approach was apparent when including limited observations and missing values. Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and ANOVA models was wider than the proposed method. b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}
